---
title: "Spatial Analysis of US Census Data in R"
subtitle: "Geometries, maps, and models"  
author: "Kyle Walker"
date: March 11, 2021
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(tigris_use_cache = TRUE)
library(tidycensus)

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

style_xaringan(
  title_slide_background_color = "#035004",
  text_color = "black",
  header_color = "#035004",
  inverse_background_color = "#035004",
  text_font_family = "Gotham Narrow",
  header_font_family = "Helvetica",
  header_font_weight = "bold",
  link_color = "#1a730f",
  code_inline_color = "#035004"
)
```

## About me

* Associate Professor of Geography at TCU 

* Spatial data science researcher and consultant

* R package developer: tidycensus, tigris, mapboxapi

* Book coming this year: _Analyzing the US Census with R_
  - These workshops are a sneak preview of the book's content!

---

## SSDAN workshop series

* Today: spatial analysis and mapping in R

* Last Thursday (March 4): an introduction to analyzing US Census data with tidycensus

* Thursday, March 25: working with US Census microdata (PUMS) with R and tidycensus

---

## Today's agenda

* Hour 1: An introduction to Census geometries and the tigris package

* Hour 2: Mapping Census data in R

* Hour 3: Spatial analysis of Census data with the sf package

---
class: middle, center, inverse

## Part 1: An introduction to Census geometries and the tigris package

---

## US Census Geography

<img src=img/census_diagram.png style="width: 500px">

.footnote[Source: [US Census Bureau](https://www2.census.gov/geo/pdfs/reference/geodiagram.pdf)]

---

