---
title: "Spatial Analysis of US Census Data in R"
subtitle: "Geometries, maps, and methods"  
author: "Kyle Walker"
date: March 11, 2021
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(tigris_use_cache = TRUE)
library(tigris)
library(tmap)
tmap_options(legend.text.size = 1)

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 12)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

style_xaringan(
  title_slide_background_color = "#035004",
  text_color = "black",
  header_color = "#035004",
  inverse_background_color = "#035004",
  text_font_family = "Gotham Narrow",
  header_font_family = "Helvetica",
  header_font_weight = "bold",
  link_color = "#1a730f",
  code_inline_color = "#035004"
)
```

## About me

* Associate Professor of Geography at TCU 

* Spatial data science researcher and consultant

* R package developer: tidycensus, tigris, mapboxapi

* Book coming this year: _Analyzing the US Census with R_
  - These workshops are a sneak preview of the book's content!

---

## SSDAN workshop series

* Today: US Census microdata and international Census data in R

* March 4: an introduction to analyzing US Census data with tidycensus ([code](https://github.com/walkerke/umich-workshop/blob/main/census-data-in-r/code/part-1-code.R) | [video recording](https://www.youtube.com/watch?v=PnFJfuJ83NI))

* March 11: spatial analysis of US Census data in R  ([code](https://github.com/walkerke/umich-workshop/blob/main/spatial-analysis/code/part-2-code.R) | [video recording](https://www.youtube.com/watch?v=GqC1HjAKui4))

---

## Today's agenda

* Hour 1: Using ACS microdata with tidycensus

* Hour 2: Analyzing and modeling microdata

* Hour 3: Bonus: international Census data resources in R

---
class: middle, center, inverse

## Part 1: Using ACS microdata with tidycensus

---

## What is "microdata?"  

{explain here}

---

## Microdata resources: DataFerrett

{discuss here}

---

## Microdata resources: bulk Census downloads

{discuss here}

---

## Microdata resources: IPUMS

{discuss here}

---

## Microdata and the Census API

{explain here how data.census.gov made it available}

---
class: middle, center, inverse

## Using microdata in tidycensus

---

## Setting up tidycensus

```{r, eval = FALSE}
library(tidycensus)

census_api_key("YOUR KEY GOES HERE", install = TRUE)
```


---

## Basic usage of `get_pums()`

{Explain the arguments here}

```{r}
library(tidycensus)

wy_pums <- get_pums(
  variables = c("PUMA", "SEX", "AGEP", "SCHL"),
  state = "WY",
  survey = "acs1",
  year = 2019
)
```

---

```{r}
wy_pums
```

---

## Understanding default data from `get_pums()`

{discuss serial number, weights, etc.}

---

## Weights and ACS microdata

```{r}
library(tidyverse)

wy_age_50 <- filter(wy_pums, AGEP == 50)

print(sum(wy_pums$PWGTP))
print(sum(wy_age_50$PWGTP))
```

---
class: middle, center, inverse

## Working with PUMS variables

---

## Variables available in the ACS PUMS

```{r, eval = FALSE}
View(pums_variables)
```

---

## Understanding PUMS variables

{info here}

---

## Recoding PUMS variables

{how-to here}

---

## Using variables filters

{show how here}

---
class: middle, center, inverse

## Public Use Microdata Areas (PUMAs)

---

## What is a PUMA?

{example from tigris here}

---

## Working with PUMAs in PUMS data

{info here}

---

## Custom queries by PUMA

{info here}

---

## Advanced example: mapping PUMS data

{calculate count, join to shapes, tmap bubbles}

---

## Part 1 exercises

{exercises here}

---
class: middle, center, inverse

## Analyzing and modeling PUMS data in R

---




